BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-07-19 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
reviewer 1 edits

Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread]. 
This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences and grearly expanded their roles in science [@ronsted2020integrative; @marsico2020small].

@@

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via commercially developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many novice botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

@@

Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota]. 
The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, beyond verifying the nations of collection it does not include data cleaning functionalities. 
While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.

@@

The formatting module has three functions. Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.

The manual review process technically only has one function that is optional and may be executed during the spatial process (after *coords2sf*), but the importance of manual review is important enough to warrant explicit mention.  
*geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/). 
Notably, many of the flags that BarnebyLives generates will be placed into columns with obviously flagged names and can be manually reviewed by the analyst, and many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.  
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.  
*label_writer* write from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
*format_database_import* will write out a spreadsheet of cleaned data in a variety of formats, currently: Jepson, Symbiota, and Consortium of Pacific Northwest herbaria are supported.

@@

The testing of the package within this manuscript was performed using a subset of the authors collections from 2018-2022, while most development was performed on their 2023 and 2024 collections. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used for most functionality. 
In total `r nrow(data)` records were used for testing various functions, these records were from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families.

@@

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives). 
The package includes three real use-case vignettes (tutorials) available on a Github Pages site (https://sagesteppe.github.io/BarnebyLives/).
The first vignette *"Preparing to use BarnebyLives!"* shows how to set up an instance for a certain geographic area (domain). 
The next two vignettes *"BarnebyLives! Running pipeline"* showcases the usage of the package for processing data entered on a spreadsheet, and *"Printing herbarium labels and exporting a digital copy of data"* how to export data in both digital and analog formats. 
*"Custom label templates"* shows how to customize labels in LaTeX, and *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium. 
All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.

>>>>

Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting Earth's biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread]. 
This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences and grearly expanded their roles in science [@ronsted2020integrative; @marsico2020small].

@@

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via commercially developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many novice botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

@@

Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota]. 
The office suite costs money, and in our experience, can be problematic to set up; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, beyond verifying the nations of collection it does not include data cleaning functionalities. 
While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfjam and pdftk, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.

@@

The formatting module has three functions. Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will supply a warning messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.

The manual review process technically only has one optional function (after *coords2sf*), however given the importance of ensuring that coordinates are appropriately located for many of the modules in the software, we suggest all users perform it.   
*geodata_writer* will write out a spatial copy of the data set to any geospatial format supported by the sf package, but defaults to writing out 'kmls' which are readily used with [Google Earth](https://earth.google.com/web/), and can also be opened in several other free geographic information system (GIS) softwares such as [QGIS](https://www.qgis.org/). 
Notably, many of the flags that BarnebyLives generates will be placed into columns with obviously flagged names and can be manually reviewed by the analyst, and many of these issues can be resolved by simply addressing the relevant issues in the original data input spreadsheet.

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the purrr package from the tidyverse.
*labels_skeleton* is an R 'script' which will require modifications to customize for institutions, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.  
*label_writer* writes from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
*format_database_import* will write out a spreadsheet of cleaned data in a variety of formats, currently: Jepson, Symbiota, and Consortium of Pacific Northwest herbaria are supported.

@@

The testing of the package within this manuscript was performed using a subset of the first author's collections from 2018-2022, while most development was performed on their 2023 and 2024 collections. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used for most functionality. 
In total `r nrow(data)` records were used for testing various functions, these records were from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families.

@@

# DATA AVAILABILITY STATEMENT
The BarnebyLives R package is open source, the development version is available on GitHub (https://github.com/sagesteppe/BarnebyLives). 
The package includes three real use-case vignettes (tutorials) available on a Github Pages site (https://sagesteppe.github.io/BarnebyLives/).
The first vignette *"Preparing to use BarnebyLives!"* shows how to set up an instance for a certain geographic area (domain). 
The next two vignettes *"BarnebyLives! Running pipeline"* showcases the usage of the package for processing data entered on a spreadsheet, and *"Printing herbarium labels and exporting a digital copy of data"* demonstrates how to export data in both digital and analog formats. 
*"Custom label templates"* shows how to customize labels in LaTeX, and *"Rendering a shipping manifest"* details how to produce a shipping manifest for gifting or transferring material to an herbarium. 
All data used in this manuscript are available at: https://github.com/sagesteppe/Barneby_Lives_dev/manuscript.
